Rising from the Ashes: A Metallicity-Dependent Star Formation Gap Splits the Milky Way's $\alpha$-Sequences

### 2025-02-14 paper/ms.tex
Spellcheck

\section{Introduction} \label{sec:intro}
% What leads to distribution of heavy elements in a galaxy
Many elements heavier than hydrogen are produced through nuclear fusion in compact objects such as supernovae, dying low mass stars, and neutron star-neutron star mergers \citep[e.g.][]{2023A&ARv..31....1A}. By necessity, stars inherit the constituitive properties of the gas from which they formed. Moreover, the surface abundance of most elements for most stars do not change over most of their lifetime. \edit1{By analyzing the surface abundances of stars, we can reconstruct the historical gas-phase composition of a galaxy.}

The enrichment of the gas-phase of a galaxy is determined by a complicated combination of physical processes - stellar evolution and supernovae, gas accretion, mergers, gas outflows from stellar and active galactic nuclei (AGN) feedback, metal mixing and diffusion, etc. Because the processes which give rise to this distribution are complex, there is almost certainly some structure in the stellar abundance distribution for every galaxy. However, it has only been definitively measured in the Milky Way, \edit1{with conflicting claims of detection \citep{2023ApJ...956L..14K} and nondetection \citep{2024IAUS..377..115N} in M31.}

@@

The stellar mass build-up of our Milky Way-like and GSE-like galaxies is given in Figure~\ref{fig:mass_size}. The upper panel shows the stellar mass history. We attempt to match the expected mass of the present-day thick disk \citep[$\sim6\times10^9\,\Msun$, horizontal blue dashed line][]{2016ARA&A..54..529B} at an evolution time of $\sim3\,\Gyr$ (corresponding to $z\sim2$, vertical dashed gray line).\footnote{Of course, this neglects the significant mass contribution of the bulge, which presumably formed earlier. However, our setup does not form a strong spheroidal component. Using the trick in e.g. \citet{2022MNRAS.515.1524Z}, we take the bulge mass to be twice the counterrotating stellar mass. At $3\,\Gyr$ in the isolated Milky Way-like galaxy, the bulge mass is $\sim7\times10^{8}\,\Msun$, or $\sim13\%$ of the total mass. The Milky Way's bulge is  $\sim1.5\times10^{10}\,\Msun$, although there is strong debate about just how much of the bulge is a classical bulge which formed before the disk \citep{2016ARA&A..54..529B}. In any case, we did not attempt to match any particular property of the bulge, though one could promote bulge formation by reducing the rotation of the gas in the inner region.} We get reasonably close at $\sim5\times10^9\,\Msun$ (blue line). For GSE, we use the best-fit mass from the $N$-body simulations of \citet{2021ApJ...923...92N} -- $5\times10^8\,\Msun$ (horizontal dashed orange line). For this, we slightly overestimate at $\sim6\times10^8\,\Msun$ (orange line).

@@

We show the bimodal and unimodal simulations' orbits in Figure~\ref{fig:orbit}. \edit1{We use a shrinking spheres center of mass method to identify the centers of the central and satellite galaxy \citep[e.g.,][]{2003MNRAS.338...14P}.\footnote{The position of the minimum potential particle in each substructured identified by \texttt{SUBFIND} is used as the starting guess, and we use an initial/final radius and step factor of $10\,\kpc$, $5\,\kpc$, and $0.9$, respectively.}} The upper and middle panels show the orbits in the $x$-$y$ and $x$-$z$ planes, respectively. The lower panel shows the separation distance as a function of time. The orbit is initially retrograde, but quickly radializes after the first pericentric passage. Coalescence occurs rapidly after the second pericentric passage at $\sim2\,\Gyr$, and \texttt{SUBFIND} ceases to recognize the satellite as a separate subhalo.

@@

\subsection{Bimodality Score}\label{ssec:bim_score}
\edit1{Given the modest size of our suite, some method for scoring the degree of bimodality for a given 1D distribution is desirable. Tests of whether a distribution is bimodal or unimodal exist -- e.g., the Hartigan dip test \citep{10.1214/aos/1176346577}, but they lack the ability to rank order based on ``bimodalitiness.'' In order to do this, we fit a given distribution of \alphaFe{} as a two-component gaussian mixture model. The bimodality score $\mathcal{B}$ is then computed as
\begin{equation}\label{eq:bimscore}
\mathcal{B} = \frac{\left|\mu_1-\mu_2\right|}{\sqrt{\sigma_1^2-\sigma_2^2}}\times w(A_2, t, k)\textrm{,}
\end{equation}
where $(\mu_1, \sigma_1)$ is the mean and standard deviation of the higher weighted mode, $(\mu_2, \sigma_2)$ likewise for the less weighted mode, $A_2$ is the amplitude of the less weighted mode, and $w$ is a penalty function with parameters $t$ and $k$ given by
\begin{equation}\label{eq:bimscore_penalty}
w(A, t, k) = \left[1+e^{-k\left(A-t\right)}\right]^{-1}\text{.}
\end{equation}
We take $t$ and $k$ to be $0.1$ and $20$, respectively.}

@@

    \gridline{\fig{bimodal_L30_frame60.pdf}{0.3\textwidth}{}
              \fig{bimodal_L30_frame70.pdf}{0.3\textwidth}{}
              \fig{bimodal_L30_frame80.pdf}{0.3\textwidth}{}
              }
    \gridline{\fig{bimodal_L30_frame90.pdf}{0.3\textwidth}{}
              \fig{bimodal_L30_frame100.pdf}{0.3\textwidth}{}
              \fig{bimodal_L30_frame110.pdf}{0.3\textwidth}{}
              }
    \gridline{\fig{bimodal_L30_frame120.pdf}{0.3\textwidth}{}
              \fig{bimodal_L30_frame200.pdf}{0.3\textwidth}{}
              \fig{bimodal_L30_frame320.pdf}{0.3\textwidth}{}
              }
  \caption{Frames from a movie showing a surface density projection of the bimodal simulation over time. In each frame, the left/right (blue/orange) column shows the gas/star surface density. The upper/lower panels show the edge-on and face-on view. Every panel is oriented with respect to the final ($t=8\,\Gyr$) snapshot. The side-length of each panel is $30\,\kpc$, and the image is a projection through a box with the same side-length. The colormap for the gas ranges from $1$ to $10^2\,\Msun/\pc^2$, while for the stars ranges from $1$ to $10^4\,\Msun/\pc^2$. \edit1{Description of movie: The disk collapses quickly, forming a disk within $500\,\Myr$. The satellite galaxy quickly passes in the background at $\sim700\,\Myr$. At $\sim2\,\Gyr$, the satellite directly merges with the central galaxy, fully coalescing by $\sim3\,\Gyr$. Over the next $5\,\Gyr$, the disk steadily grows, expanding in size.} (A full movie is \edit1{linked in submission}.)}
  \label{fig:projections}
\end{figure*}

@@

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{alpha_vs_tform_obserr.pdf}
  \caption{\edit1{In the bimodal simulation, a gap appears when plotting stars in a $0.05\,\dex$ bin at $\FeH=0$ on the \MgFe{}-$t_{\textrm{form}}$ plane (blue, left panel). In the unimodal simulation there is no such gap (orange, left panel). In the middle panel, we show the same data but assuming errors in \FeH{}, \MgFe{}, and $t_{\textrm{form}}$ of $0.0075\,\dex$, $0.012\,\dex$, and $1\,\Gyr$, respectively. The gap in the bimodal simulation appears weakly as two populations overlapping in age. In the right panel we show stars in the Milky Way from APOKASC3 in $0.2\,\dex$ bins at $\FeH=-0.5$ (blue) and $0$ (orange). The low-metallicity bin is where the Milky Way bimodality is strongest while there is no bimodality at the solar-metallicty bin (see Figure~\ref{fig:fig1}). Offsets have been added to \MgFe{} values for clarity: $-0.3$ in the unimodal points in the left and middle panels, and $+0.1$ and $-0.1$ in the low/solar-metallicity bins, respectively, in the right panel.}}
  \label{fig:alpha_vs_tform}
\end{figure*}

@@

\edit1{The right panel shows observational data from APOKASC-3. We show in the blue, upper data a $0.2\,\dex$ bin centered at $\FeH=-0.5$ and in the orange, lower data centered at $\FeH=0$. In the Milky Way, there is a bimodal and unimodal \MgFe{} distribution at these metallicities, respectively. At $\FeH=-0.5$, one could weakly argue that there are separate populations as in the simulation. However, the large uncertainties at the relevant ages ($\sim1\,\Gyr$ at ages of $\sim8\,\Gyr$) and the small sample size (125 and 1175 at $\FeH=-0.5$ and $0$, respectively) prevents a definitive statement from being made. Larger sample sizes and more precise age estimates in the future would clarify the connection between simulation and observation, although achieving age uncertainities of $<10\%$ for such old stars is very challenging \citep[e.g.][]{2010ARA&A..48..581S}.}

@@

\edit1{Simulations from r onwards, which have $\mathcal{B}>2.25$, are considered bimodal. This demarcation was chosen by eye using Figure~\ref{fig:all_hist}. For these simulations, we plot the trough \MgFe{} (the location of the minimum between the two maxima) as a vertical line. This value was determined by taking the location of the minimum of the distribution between $0.25$ and $0.35\,\dex$, and is also given in Table~\ref{tab:my-table}.}

@@

\edit1{Most of the bimodal simulations (r through aa) have a gap in the distribution at around the time of the merger ($\sim2$-$3\,\Gyr$, depending on the orbital parameters). Before this gap, star formation occurs above the trough while after it occurs below the trough. However, there are two exceptions: simulation~t does not have a full gap (though the density does decrease), and simulation~w is irregular, with star formation simultaneously occuring above and below the trough.}

\edit1{In the unimodal simulations (a through q), there does appear to be gaps (d, f, g, h, and o). However, in these cases, there is not a large amount of star formation occuring before the gap, and so the high-$\alpha$ sequence is underemphasized and does not lead to a significant bimodality.}

@@

\edit1{This perspective is bolstered by examining the distributions of \MgFe{} at $\FeH=0$ for the full simulation suite (Figure~\ref{fig:all_hist}). We order these simulations alphabetically using the bimodality score ($\mathcal{B}$, see Section~\ref{ssec:bim_score}). Simulations~r through aa are considered bimodal based on a visual inspection, and the location of their trough \MgFe{} (minimium of the distribution) is indicated with a vertical line.}

@@

\subsection{Connection to High Redshift Quenching}\label{ssec:obshiz}
\edit1{One plausible avenue to producing a \FeH{}-dependent halt in star formation is through a global quiescent period. Galaxies which undergo a starburst to quiescence to rejuvenation sequence (post-starburst galaxies, or PSBs) are observed at high-$z$ ($z>1$), and may be plausible Milky Way-progenitors.} With abundance matching, we expect the Milky Way's total stellar mass to be $\sim10^{10.3}\,\Msun$ at $z\sim2$ \citep{2013ApJ...771L..35V}. A number of authors have explored PSBs and quiescent galaxies at slightly higher masses at $z\sim2$, with large advances in the post-JWST era.

@@

\citet{2018MNRAS.477.5072M} found that Milky Way-like bimodalities are rare in EAGLE, occuring in $\sim5\%$ of galaxies. \citet{2021MNRAS.501..236D,2022MNRAS.515.1430D} showed that merger-induced quenching in zooms can occur in the EAGLE model \citep[see also][]{2017MNRAS.465..547P}. However, the situation may be different in the lower resolution large box. Furthermore, if the proposed starburst-quenching phase is driven by AGN feedback, then the outcome of any particular cosmological simulation with regards to the bimodality is intimately tied to its AGN model. Unfortunately, such models are highly uncertain \citep[e.g.][]{2022MNRAS.511.3751H}.

@@

\edit1{Another natural next step would be to extend the idealized simulations in this work to a wider range of orbits, galaxy properties, and feedback models. However, several aspects of our setup are unrealistic, for example, (1) the simulation is not in an expanding universe, (2) our feedback model is weaker than typical ones calibrated to full cosmological box simulations, (3) the initial conditions have a steeper potential well prior to star formation than in the real universe, and (4) our halos lack small-scale substructure. The simplicity of our setup aids interpretation, but also limits it applicability to the real unvierse. Something along the lines of the genetic modification technique to explore various mergers as done in this work may be useful \citep{2016MNRAS.455..974R,2017MNRAS.465..547P}.}

>>>>

\section{Introduction} \label{sec:intro}
% What leads to distribution of heavy elements in a galaxy
Many elements heavier than hydrogen are produced through nuclear fusion in compact objects such as supernovae, dying low mass stars, and neutron star-neutron star mergers \citep[e.g.][]{2023A&ARv..31....1A}. By necessity, stars inherit the constitutive properties of the gas from which they formed. Moreover, the surface abundance of most elements for most stars do not change over most of their lifetime. \edit1{By analyzing the surface abundances of stars, we can reconstruct the historical gas-phase composition of a galaxy.}

The enrichment of the gas-phase of a galaxy is determined by a complicated combination of physical processes - stellar evolution and supernovae, gas accretion, mergers, gas outflows from stellar and active galactic nuclei (AGN) feedback, metal mixing and diffusion, etc. Because the processes which give rise to this distribution are complex, there is almost certainly some structure in the stellar abundance distribution for every galaxy. However, it has only been definitively measured in the Milky Way, \edit1{with conflicting claims of detection \citep{2023ApJ...956L..14K} and non-detection \citep{2024IAUS..377..115N} in M31.}

@@

The stellar mass build-up of our Milky Way-like and GSE-like galaxies is given in Figure~\ref{fig:mass_size}. The upper panel shows the stellar mass history. We attempt to match the expected mass of the present-day thick disk \citep[$\sim6\times10^9\,\Msun$, horizontal blue dashed line][]{2016ARA&A..54..529B} at an evolution time of $\sim3\,\Gyr$ (corresponding to $z\sim2$, vertical dashed gray line).\footnote{Of course, this neglects the significant mass contribution of the bulge, which presumably formed earlier. However, our setup does not form a strong spheroidal component. Using the trick in e.g. \citet{2022MNRAS.515.1524Z}, we take the bulge mass to be twice the counter-rotating stellar mass. At $3\,\Gyr$ in the isolated Milky Way-like galaxy, the bulge mass is $\sim7\times10^{8}\,\Msun$, or $\sim13\%$ of the total mass. The Milky Way's bulge is  $\sim1.5\times10^{10}\,\Msun$, although there is strong debate about just how much of the bulge is a classical bulge which formed before the disk \citep{2016ARA&A..54..529B}. In any case, we did not attempt to match any particular property of the bulge, though one could promote bulge formation by reducing the rotation of the gas in the inner region.} We get reasonably close at $\sim5\times10^9\,\Msun$ (blue line). For GSE, we use the best-fit mass from the $N$-body simulations of \citet{2021ApJ...923...92N} -- $5\times10^8\,\Msun$ (horizontal dashed orange line). For this, we slightly overestimate at $\sim6\times10^8\,\Msun$ (orange line).

@@

We show the bimodal and unimodal simulations' orbits in Figure~\ref{fig:orbit}. \edit1{We use a shrinking spheres center of mass method to identify the centers of the central and satellite galaxy \citep[e.g.,][]{2003MNRAS.338...14P}.\footnote{The position of the minimum potential particle in each substructure identified by \texttt{SUBFIND} is used as the starting guess, and we use an initial/final radius and step factor of $10\,\kpc$, $5\,\kpc$, and $0.9$, respectively.}} The upper and middle panels show the orbits in the $x$-$y$ and $x$-$z$ planes, respectively. The lower panel shows the separation distance as a function of time. The orbit is initially retrograde, but quickly radializes after the first pericentric passage. Coalescence occurs rapidly after the second pericentric passage at $\sim2\,\Gyr$, and \texttt{SUBFIND} ceases to recognize the satellite as a separate subhalo.

@@

\subsection{Bimodality Score}\label{ssec:bim_score}
\edit1{Given the modest size of our suite, some method for scoring the degree of bimodality for a given 1D distribution is desirable. Tests of whether a distribution is bimodal or unimodal exist -- e.g., the Hartigan dip test \citep{10.1214/aos/1176346577}, but they lack the ability to rank order based on ``bimodalitiness.'' In order to do this, we fit a given distribution of \alphaFe{} as a two-component Gaussian mixture model. The bimodality score $\mathcal{B}$ is then computed as
\begin{equation}\label{eq:bimscore}
\mathcal{B} = \frac{\left|\mu_1-\mu_2\right|}{\sqrt{\sigma_1^2-\sigma_2^2}}\times w(A_2, t, k)\textrm{,}
\end{equation}
where $(\mu_1, \sigma_1)$ is the mean and standard deviation of the higher weighted mode, $(\mu_2, \sigma_2)$ likewise for the less weighted mode, $A_2$ is the amplitude of the less weighted mode, and $w$ is a penalty function with parameters $t$ and $k$ given by
\begin{equation}\label{eq:bimscore_penalty}
w(A, t, k) = \left[1+e^{-k\left(A-t\right)}\right]^{-1}\text{.}
\end{equation}
We take $t$ and $k$ to be $0.1$ and $20$, respectively.}

@@

    \gridline{\fig{bimodal_L30_frame60.pdf}{0.3\textwidth}{}
              \fig{bimodal_L30_frame70.pdf}{0.3\textwidth}{}
              \fig{bimodal_L30_frame80.pdf}{0.3\textwidth}{}
              }
    \gridline{\fig{bimodal_L30_frame90.pdf}{0.3\textwidth}{}
              \fig{bimodal_L30_frame100.pdf}{0.3\textwidth}{}
              \fig{bimodal_L30_frame110.pdf}{0.3\textwidth}{}
              }
    \gridline{\fig{bimodal_L30_frame120.pdf}{0.3\textwidth}{}
              \fig{bimodal_L30_frame200.pdf}{0.3\textwidth}{}
              \fig{bimodal_L30_frame320.pdf}{0.3\textwidth}{}
              }
  \caption{Frames from a movie showing a surface density projection of the bimodal simulation over time. In each frame, the left/right (blue/orange) column shows the gas/star surface density. The upper/lower panels show the edge-on and face-on view. Every panel is oriented with respect to the final ($t=8\,\Gyr$) snapshot. The side-length of each panel is $30\,\kpc$, and the image is a projection through a box with the same side-length. The color map for the gas ranges from $1$ to $10^2\,\Msun/\pc^2$, while for the stars ranges from $1$ to $10^4\,\Msun/\pc^2$. \edit1{Description of movie: The disk collapses quickly, forming a disk within $500\,\Myr$. The satellite galaxy quickly passes in the background at $\sim700\,\Myr$. At $\sim2\,\Gyr$, the satellite directly merges with the central galaxy, fully coalescing by $\sim3\,\Gyr$. Over the next $5\,\Gyr$, the disk steadily grows, expanding in size.} (A full movie is \edit1{linked in submission}.)}
  \label{fig:projections}
\end{figure*}

@@

\begin{figure*}
  \centering
  \includegraphics[width=\textwidth]{alpha_vs_tform_obserr.pdf}
  \caption{\edit1{In the bimodal simulation, a gap appears when plotting stars in a $0.05\,\dex$ bin at $\FeH=0$ on the \MgFe{}-$t_{\textrm{form}}$ plane (blue, left panel). In the unimodal simulation there is no such gap (orange, left panel). In the middle panel, we show the same data but assuming errors in \FeH{}, \MgFe{}, and $t_{\textrm{form}}$ of $0.0075\,\dex$, $0.012\,\dex$, and $1\,\Gyr$, respectively. The gap in the bimodal simulation appears weakly as two populations overlapping in age. In the right panel we show stars in the Milky Way from APOKASC3 in $0.2\,\dex$ bins at $\FeH=-0.5$ (blue) and $0$ (orange). The low-metallicity bin is where the Milky Way bimodality is strongest while there is no bimodality at the solar-metallicity bin (see Figure~\ref{fig:fig1}). Offsets have been added to \MgFe{} values for clarity: $-0.3$ in the unimodal points in the left and middle panels, and $+0.1$ and $-0.1$ in the low/solar-metallicity bins, respectively, in the right panel.}}
  \label{fig:alpha_vs_tform}
\end{figure*}

@@

\edit1{The right panel shows observational data from APOKASC-3. We show in the blue, upper data a $0.2\,\dex$ bin centered at $\FeH=-0.5$ and in the orange, lower data centered at $\FeH=0$. In the Milky Way, there is a bimodal and unimodal \MgFe{} distribution at these metallicities, respectively. At $\FeH=-0.5$, one could weakly argue that there are separate populations as in the simulation. However, the large uncertainties at the relevant ages ($\sim1\,\Gyr$ at ages of $\sim8\,\Gyr$) and the small sample size (125 and 1175 at $\FeH=-0.5$ and $0$, respectively) prevents a definitive statement from being made. Larger sample sizes and more precise age estimates in the future would clarify the connection between simulation and observation, although achieving age uncertainties of $<10\%$ for such old stars is very challenging \citep[e.g.][]{2010ARA&A..48..581S}.}

@@

\edit1{Simulations from r onward, which have $\mathcal{B}>2.25$, are considered bimodal. This demarcation was chosen by eye using Figure~\ref{fig:all_hist}. For these simulations, we plot the trough \MgFe{} (the location of the minimum between the two maxima) as a vertical line. This value was determined by taking the location of the minimum of the distribution between $0.25$ and $0.35\,\dex$, and is also given in Table~\ref{tab:my-table}.}

@@

\edit1{Most of the bimodal simulations (r through aa) have a gap in the distribution at around the time of the merger ($\sim2$-$3\,\Gyr$, depending on the orbital parameters). Before this gap, star formation occurs above the trough while after it occurs below the trough. However, there are two exceptions: simulation~t does not have a full gap (though the density does decrease), and simulation~w is irregular, with star formation simultaneously occurring above and below the trough.}

\edit1{In the unimodal simulations (a through q), there does appear to be gaps (d, f, g, h, and o). However, in these cases, there is not a large amount of star formation occurring before the gap, and so the high-$\alpha$ sequence is underemphasized and does not lead to a significant bimodality.}

@@

\edit1{This perspective is bolstered by examining the distributions of \MgFe{} at $\FeH=0$ for the full simulation suite (Figure~\ref{fig:all_hist}). We order these simulations alphabetically using the bimodality score ($\mathcal{B}$, see Section~\ref{ssec:bim_score}). Simulations~r through aa are considered bimodal based on a visual inspection, and the location of their trough \MgFe{} (minimum of the distribution) is indicated with a vertical line.}

@@

\subsection{Connection to High Redshift Quenching}\label{ssec:obshiz}
\edit1{One plausible avenue to producing a \FeH{}-dependent halt in star formation is through a global quiescent period (see Appendix~\ref{app:all_sfh}). Galaxies which undergo a starburst to quiescence to rejuvenation sequence (post-starburst galaxies, or PSBs) are observed at high-$z$ ($z>1$), and may be plausible Milky Way-progenitors.} With abundance matching, we expect the Milky Way's total stellar mass to be $\sim10^{10.3}\,\Msun$ at $z\sim2$ \citep{2013ApJ...771L..35V}. A number of authors have explored PSBs and quiescent galaxies at slightly higher masses at $z\sim2$, with large advances in the post-JWST era.

@@

\citet{2018MNRAS.477.5072M} found that Milky Way-like bimodalities are rare in EAGLE, occurring in $\sim5\%$ of galaxies. \citet{2021MNRAS.501..236D,2022MNRAS.515.1430D} showed that merger-induced quenching in zooms can occur in the EAGLE model \citep[see also][]{2017MNRAS.465..547P}. However, the situation may be different in the lower resolution large box. Furthermore, if the proposed starburst-quenching phase is driven by AGN feedback, then the outcome of any particular cosmological simulation with regards to the bimodality is intimately tied to its AGN model. Unfortunately, such models are highly uncertain \citep[e.g.][]{2022MNRAS.511.3751H}.

@@

\edit1{Another natural next step would be to extend the idealized simulations in this work to a wider range of orbits, galaxy properties, and feedback models. However, several aspects of our setup are unrealistic, for example, (1) the simulation is not in an expanding universe, (2) our feedback model is weaker than typical ones calibrated to full cosmological box simulations, (3) the initial conditions have a steeper potential well prior to star formation than in the real universe, and (4) our halos lack small-scale substructure. The simplicity of our setup aids interpretation, but also limits it applicability to the real universe. Something along the lines of the genetic modification technique to explore various mergers as done in this work may be useful \citep{2016MNRAS.455..974R,2017MNRAS.465..547P}.}
